---
title: "5240 Workshop 05"
author: "Digvijay Jondhale 0862899"
format: pdf
---

### Assuming we have no other information about the older brother (in other words, he can be treated as a randomly selected 8th grader), what is the probability that the older brother will run the 5km in less than 24 minutes?

**Ans : **
Given information : 

* For Older Brother (Grade 8): 
    * $\mu$ = 25
    * $\sigma$ = 2 
    * $X$ = 24
    
To find the probability that Older brother run's the 5 km race in less than 24 mins, we can calculate Z score for x = 24 and then find the probability P(z<calculated z value)
```{r}
z_older_brother <- scale(24, center = 25, scale = 2)
round(100*(pnorm(z_older_brother)),2)
```
**Probability that the older brother will run the 5km in less than 24 minutes is 30.85% or 0.3085**

\newpage
### Suppose that the older brother runs his race in 23 minutes and 30 seconds and the younger brother runs his race in 27 minutes. Using Z-scores, determine which brother ran his race faster, relative to their respective grades.

**Ans : ** To determine this we can calculate z score for both the brothers , as z scores allow us to compare different normal distributions on same scale.
the lower z score will correspond to the brother who ran his race faster.

Given information from the table : 

  * For Older Brother (Grade 8): 
    * $\mu$ = 25
    * $\sigma$ = 2 
    * $X$ = 23.5

  * For Younger Brother (Grade 6): 
    * $\mu$ = 28
    * $\sigma$ = 3 
    * $X$ = 27
    
```{r}
z_older <- scale(23.5, center = 25, scale = 2)
z_younger <- scale(27, center = 28, scale = 3)

z_older
```
```{r}
z_younger
```
Z-score for older brother (-0.75) < Z-score for younger brother (-0.33),
Hence **Older brother** ran his race faster.

