---
title: "5240 Workshop Exercise 02"
author: "Digvijay Jondhale 0862899"
format: pdf
---

## Loading the Football and Weights of teams dataset 

```{r}
football_data <- read.csv(file= "ws02-exercise-football.csv", header=TRUE)
weight_of_teams <- read.csv(file = "football_weights.csv", header=TRUE)

str(football_data)
str(weight_of_teams)

```

## Visualization's

```{r}


boxplot(x=list(weight_of_teams$broncos,weight_of_teams$cowboys,weight_of_teams$dolphins,weight_of_teams$fortyniners,weight_of_teams$packers),
        xlab = "Football Teams (American)",
        ylab = "Weight of Footbal Players (in pounds)",
        names = c("Broncos","Cowboys","Dolphins","Fortyniners","Packers"),
        col = c("#780000","#c1121f","#fdf0d5","#003049","#669bbc"),
        main = "Weight of Players Vs Football Teams"
        )

```

## Analysis from the boxplot 

1. Dolphis have higher variablity of weights in their players as the IQR has larger height meaning 50% of the players in Dolphis have larger spread in their weights and also the whiskers are spread out.

2. Broncos and Cowboys have similar median, but few players in Cowboys are underweight as indicated by the outliers . On the other hand , Cowboys's players weights is more uniform because the box size (IQR) has short height meaning that 50% pf the data is centered aroud the median.

3. Fortyniners have variety of players , some players below 220 and some as high as 270., also the larger box size indicates that Fortyniners also have variation in  large chunk of players.

4. Packers have similar spread of weights like Fortyniners , but they are less skewed as compared to Fortyniners as the median is nearly in center of the box.