---
title: "5240 Workshop 09"
author: "Digvijay Jondhale 0862899"
format: pdf
---

### Loading the Data ...
```{r}
library(tidyr)
library(ggplot2)
football <- read.csv(file = 'footballweights.csv')
```

```{r}
str(football)

```
### a) Converting to longer format 
```{r}

long_football <- pivot_longer(
  football,
  cols = everything(),
  names_to = "team",
  values_to = "weights"
)

```
```{r}
str(long_football)
```

## b) Box Plot
```{r}

ggplot(long_football, aes(x = team, y = weights)) +
  geom_boxplot(fill = c("#780000","#c1121f","#fdf0d5","#003049","#669bbc"), col="black") + 
  labs(
    title = "Team Player's Weights Distribution",
    x = "Team",
    y = "Weight"
  ) +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
```

## c) ANOVA 

*Ho : * All Teams have same mean weights of players.
*Ha : * At least one Team has different mean weight of players.
```{r}
anova_result <- aov(weights~team, data = long_football)
summary(anova_result)
```
As the p-values(0.03) from the anova is < 0.05, we have to reject the null hypothesis and conclude that at least one team has a different mean weight.
To find the teams who has different mean weight,we can perform Tukey's Honestly Significant Difference Test.

## d) Tukey's Honestly Significant Difference Test
```{r}
TukeyHSD(anova_result)
```
### Results 
Fortyniners and cowboys are the teams , who has significant difference in their mean weights with fortyniners mean score significantly lower than cowboys by 16.17 units.
