---
title: "5240 Workshop 07"
author: "Digvijay Jondhale 0862899"
format: pdf
---

```{r}
library(dplyr)
```

### Loading the results ...
```{r}
results <- readRDS(file="5240-mean-reject-ws07.rds")
glimpse(results)
```
### Hypothesis 

**Ho (Null) :** Proportions of simulation rejected using CLT based approach was 10% (p = 0.10).

**Ha (Alternate) :** Proportion of simulation rejected using CLT based approach was not equal to 10% (p != 0.10)

$\alpha$ = 0.05


### P value 

#### Boot Method
```{r}
boot_sum <- sum(results$boot)
boot_sum
```
```{r}
binom.test(boot_sum,10000, 0.10)
```
 
**p value (boot method) : ** 3.03e-06



#### CLT Method
```{r}
clt_sum <- sum(results$clt)
clt_sum
```
```{r}
binom.test(clt_sum,10000,0.10)
```
**p value (clt method) : **  < 2.2e-16



#### Exact Method
```{r}
exact_sum <- sum(results$exact)
exact_sum

binom.test(exact_sum,10000,0.10)
```
**p value (exact method) : **  0.3681

### Conclusion 

For **clt based approach** , the p-value was less than 2.2e-16, which is extremely less than significance level 0.05, providing a strong evidence to reject the null hypothesis. Hence proportion of simulation rejected using clt based approach was different than 10%.

Similar result was observed for **boot method** (p-value(3.03e-06) < 0.05), rejecting null hypothesis, and making a conclusion that proportion of simulation rejected using boot based method was different from 10%.


p - value for  **exact method** (p-value(0.3681) > 0.05), accepting null hypothesis, and making a conclusion that proportion of simulation rejected using boot based method was equal to 10%.

