---
title: "Workshop 08"
author: "Digvijay Jondhale"
format: pdf
---

## Problem Statement

The GitHub repository for the Tidy Tuesday project contains a dataset for each Tuesday from the past ~6 years; i.e., around 300 datasets. A random sample of 40 dataset names was taken from the full list and the number of characters in each title was found. These values are contained in the file, ws08-exercise-title_length.rds.

```{r}

title_data <- readRDS("ws08-exercise-title_length.rds")

summary(title_data)
length(title_data) 
```

## To check for normality conditions of clt , the following conditions should be satisfied : 

 * Random sampling 
 * Sample size (n>=30)
 * Independence
 * Plotting rhe distribution and checking its shape
 
### Random sampling :
This condition is met , as it is already stated in the problem statement that the sample of 40 datasets was randomly sampled.

### Sample size
The sample size for clt to apply should be minimum 30, as the sample size is 40 >30 , this condition has been met.
```{r}
if (length(title_data) >= 30) {
  cat("Sample size is large enough (n >= 30) for CLT to apply.\n")
} else {
  cat("Sample size is not large enough (n < 30)for clt to apply .\n")
      }
```


### Independence 
For independence , we need to make sure that the sample size n = 40 should be less than 10% of the population size (300).
```{r}
population <- 300
independence <- length(title_data) < (0.1 * population)
cat("Independence Condition Met:", independence, "\n")
```
Form the above results ,the conditions for independence are not met.

### Checking distribution shape 
```{r}
hist(title_data, main = "Distribution of Dataset Title Lengths",
     xlab = "Title Length (Number of Characters)", col = "skyblue", border = "white")
```
From the histogram above , it is clear that the shape is unimodal, but it is right skewed and is not normally distributed. Hence CLT does not applies here as independence condition was not met.

### Results can be verified by Shapiro-Wilk test
```{r}
shapiro_test <- shapiro.test(title_data)
cat("Shapiro-Wilk p-value:", shapiro_test$p.value, "\n")
if (shapiro_test$p.value < 0.05) {
  cat("Not Normally Distributed (p < 0.05).\n")
} else {
  cat("Approximately Normally Distributed (p >= 0.05).\n")
}



```

### Results
The above test results confirm our analysis, that Central Limit theorem does not apply as the independence condition was violated, indicating that the sample means are **Not approximately Normally** distributed.

Note : 
*As the sample size = 40 was > 30 ,If we overlook the 10% rule of sample size then we can get a approximately Normal distribution of means using CLT*
